Abstractions for Multi-Sorted Substitutions

### 2023-09-19 paper/paper.tex
Update.

  We identify two drawbacks with this style of syntax:
  \begin{enumerate}
  \item
    the syntactic categories (\ACode{\ACon{Kind}},
    \ACode{\ACon{Type}}, and \ACode{\ACon{Expr}})
    have different types, which makes it difficult to treat them uniformly; and
  \item the different sorts of variables are modeled separately, which requires to define not just
    type-in-type and expression-in-expression substitution, but also type-in-expression substitution.
    Consequently, interaction lemmas between those substitutions are required.
    In general, this leads to a combinatory explosion of
    substitutions, e.g.\ adding kind variables would lead to a total of 6
    different substitutions and corresponding interaction lemmas.
  \end{enumerate}
  To avoid these drawbacks, we instead use a multi-sorted syntax.

@@

  Note that there is no straightforward way to construct a
  multi-sorted syntax with intrinsic typing: in a direct translation,
  the type of terms \ACode{\ACon{\_⊢\_}} would be indexed by itself, which
  most proof assistants forbid to avoid breaking logical consistency.
  We conjecture that an encoding with universe codes is possible, as
  it has been done for intrinsically typed languages with dependent
  types, but those constructions come with their own drawbacks and complexity.

@@

  \subsection{A Structure for Multi-Sorted Syntax}
  \label{sec:syntax:structure}
  The multi-sorted syntax makes it easy to define a structure for
  syntaxes of arbitrary object-languages, i.e.\ syntaxes with
  arbitrary amounts of syntactic categories and variables:
  \LibCode\KSyntax
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \ACode{\AField{`\_}} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FSyntaxInst

@@

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

  We also specify notation, which makes it less cumbersome
  to explicitly pass a particular Kit to those definitions:
  \begin{itemize}
  \item \ACode{S \ADef{∋/⊢[} K \ADef{]} s} is \ACode{S ∋/⊢ s} for some \ACode{\ACon{Kit}} \ACode{K}; and
  \item \ACode{S \ADef{–[} K \ADef{]→} s} is \ACode{S \ADef{→ₖ} s} for some \ACode{\ACon{Kit}} \ACode{K}.
  \end{itemize}

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  The fields of this structure are:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result from \ACode{S₂ ∋/⊢ S}
    to \ACode{S₂ ⊢ S} via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

@@

  We start with the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}} to it. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    Hence, the use of \ACode{`/id}.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm to account for the bound variable.
  \end{itemize}

  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

  \ExampleCode\FExampleTrav

>>>>

  We identify two drawbacks with this style of syntax:
  \begin{enumerate}
  \item
    the syntactic categories (\ACode{\ACon{Kind}},
    \ACode{\ACon{Type}}, and \ACode{\ACon{Expr}})
    have different types, which makes it difficult to treat them uniformly; and
  \item the different sorts of variables are modeled separately, which requires to define not just
    type-in-type and expression-in-expression substitution, but also type-in-expression substitution.
    Consequently, interaction lemmas between those substitutions are required.
    % In general, this leads to a combinatory explosion of
    % substitutions, e.g.\ adding kind variables would lead to a total of 6
    % different substitutions and corresponding interaction lemmas.
  \end{enumerate}
  To avoid these drawbacks, we instead use a multi-sorted syntax.

@@

  Note that there is no straightforward way to construct a
  multi-sorted syntax with intrinsic typing: in a direct translation,
  the type of terms \ACode{\ACon{\_⊢\_}} would be indexed by itself, which
  most proof assistants forbid to avoid breaking logical consistency.
  We conjecture that an encoding with universe codes is possible, as
  it has been done for intrinsically typed languages with dependent
  types\cite{TODO}, but those constructions come with additional
  complexity and are less suitable to automation.

@@

  \subsection{A Structure for Multi-Sorted Syntax}
  \label{sec:syntax:structure}
  The regularity of the multi-sorted syntax makes it easy to define a
  structure for syntaxes of arbitrary object-languages, i.e.\ syntaxes
  with arbitrary amounts of syntactic categories and variables:
  \LibCode\KSyntax
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \ACode{\AField{`\_}} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FSyntaxInst

@@

  Building on the operations and axioms of a \ACode{\ACon{Kit}}, we
  derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the \ACode{\ACon{Kit}} provides the
    necessary \ACode{\AField{wk}} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACode{\ACon{zero}} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \ACode{\AField{wk}} on terms:
    \LibCode\KWeaken
  \item
    Maps have an extensional equality:
    \LibCode\KEq
  \item
    For simplicity we postulate functional extensionality:\footnote{
      The actual implementation does \emph{not} use any postulates.
      The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
    \LibCode\KFunExt
  \item
    Lifting the identity map is a (larger) identity map:
    \LibCode\KIdLift
    % \LibCode\KIdLiftProof
  \end{itemize}

  We also specify notation, which makes it less cumbersome
  to explicitly pass a particular \ACode{\ACon{Kit}} to those definitions:
  \begin{itemize}
  \item \ACode{S \ADef{∋/⊢[} K \ADef{]} s} is \ACode{S ∋/⊢ s} for some \ACode{\ACon{Kit}} \ACode{K}; and
  \item \ACode{S \ADef{–[} K \ADef{]→} s} is \ACode{S \ADef{→ₖ} s} for some \ACode{\ACon{Kit}} \ACode{K}.
  \end{itemize}

  The operation of applying a map to a term depends on the concrete
  structure of a term, so we define a structure for it:
  \LibCode\KTraversal
  The fields of this structure have the following meaning:
  \begin{itemize}
  \item
    \ACode{t \AField{⋯} ϕ} applies the map \ACode{ϕ} (a renaming or
    substitution) to the term \ACode{t}.
  \item
    \ACode{\AField{⋯-var}} states that applying a map \ACode{ϕ} to a variable
    term\\ \ACode{\AField{`} x}, is the same as applying \ACode{ϕ} to
    the variable \ACode{x}, and then converting the result to a term
    via \ACode{\AField{id/`}}.
  \item
    \ACode{\AField{⋯-id}} states that applying the identity map
    \ACode{id} to a term does not change the term.
  \end{itemize}

@@

  We start with the operation of applying a map to a term:
  \AppCode\FTraversalOp
  The interesting cases are those with variables and binders:
  \begin{itemize}
  \item 
    In the variable case \ACode{(\ACon{`} x) \ADef{⋯} ϕ}, we first apply the map
    \ACode{ϕ} to the variable \ACode{x}. If \ACode{ϕ} is a renaming,
    we get back a variable and need to apply the variable constructor
    \ACode{\ACon{`\_}} to it. If \ACode{ϕ} is a substitution,
    we get back a term that we can use directly.
    This is exactly what \ACode{\AField{`/id}} does.
  \item 
    In cases where the operation needs to go under a binder, like
    \ACode{(\ACon{λx} e) \ADef{⋯} ϕ}, we lift the map using \AField{\_↑\_} before
    we apply it to the subterm to account for the bound variable.
  \end{itemize}

  Then we prove that applying an identity map does not change the term:
  \AppCode\FTraversalId
  \AppCode\FTraversalIdProofInteresting
  The interesting cases are:
  \begin{itemize}
  \item
    the variable case \ACode{\ACon{`} x}, where we need to combine the
    \ACon{\AField{id/`}} from the identity map, with the \ACon{\AField{`/id}}
    from the traversal operation using \ACon{\AField{`/`-is-`}}.
  \item
    the lambda abstraction case \ACode{\ACon{λx} e}, where the
    traversal lifts the identity over its binder. Here
    \ACon{\AField{id↑-id}} is used to state that this lifted identity
    map is still the identity.
  \item
    all other cases are analogous and hence omitted.
  \end{itemize}

  Finally, we instanciate the \ACode{\ACon{Traversal}} structure:
  % \LibCode\FTraversalIdProofRest
  \AppCode\FTraversal

  % \ExampleCode\FExampleTrav
